The Evolution of Phenotypically Invariant Genetic Networks

### 2018-03-07 kalman_walk_paper.tex
reordered statment of main purpose

It is an overarching goal of many biological subdisciplines 
to attain a general understanding of the function and evolution of the 
complex molecular machinery that translates an organism's genome 
into the characteristics on which natural selection acts,
the phenotype.
% Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
%Attaining a general understanding of the functioning and evolution of this molecular machinery
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus has the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.
%The dynamics of the molecular machinery and its interactions with the environment
%can be mathematically described as a dynamical system \citep{jaeger2015comet}.
%Movement in this direction is ongoing, as 
There is a substantial amount of work studying the evolution of gene regulatory networks, in both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
and empirically inspired computational
\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico} frameworks.
% If we allow the reasonable assumption that the genotype-phenotype map 
% can be represented as a system of differential equations \citep{jaeger2015comet}, 
% we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

It is well known that in many contexts mathematical models
can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modelled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments on the ``black box'' are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases in which,
despite the existence of a theoretical one-to-one mapping between a model and behavior,
tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
%even if a one-to-one mapping between model and behavior is possible in theory,
%even tiny amounts of noise can make inference problems nonidentifiable in practice \citep{petrov2005well}.

@@

In this paper we outline a theoretical framework to study the evolution of biological systems, such as gene regulatory networks.
We study the evolution of an optimally adapted population under stable conditions, that is where phenotype is under stabilizing selection.
Even if the phenotype remains stable over evolutionary time, the underlying mechanism might not remain so,
as many distinct (and mutationally connected) molecular pathways can realize identical phenotypes.
Below, we first apply results from system theory which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes.
Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
system drift may result in reproductive incompatibility between sufficiently isolated populations,
%isolated for a sufficiently long period of time, 
even in the absence of any sort of adaptive, selective, or environmental change. 
We then use quantitative genetic theory to estimate how quickly 
reproductive incompatibility due to system drift manifests.

>>>>

It is an overarching goal of many biological subdisciplines 
to attain a general understanding of the function and evolution of the 
complex molecular machinery that translates an organism's genome 
into the characteristics on which natural selection acts,
the phenotype.
% Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
%Attaining a general understanding of the functioning and evolution of this molecular machinery
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus has the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.
%The dynamics of the molecular machinery and its interactions with the environment
%can be mathematically described as a dynamical system \citep{jaeger2015comet}.
%Movement in this direction is ongoing, as 
There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness} 
and empirically inspired
\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.
% If we allow the reasonable assumption that the genotype-phenotype map 
% can be represented as a system of differential equations \citep{jaeger2015comet}, 
% we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

It is well known that in many contexts mathematical models
can fundamentally be \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modelled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The inherent nonidentifiability of chemical reaction networks is sometimes referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases in which,
despite the existence of a theoretical one-to-one mapping between a model and behavior,
tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.
%even if a one-to-one mapping between model and behavior is possible in theory,
%even tiny amounts of noise can make inference problems nonidentifiable in practice \citep{petrov2005well}.

@@

The main purpose of this paper is to explore this general idea
in the concrete framework of linear systems theory,
thought of as modeling gene regulatory networks.
First, we apply results from system theory which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes.
This suggests that, quite generally, there are many directions in which a gene network can drift
without changing the phenotype.
Since these phenotypically equivalent gene networks are not necessarily compatible with one another, 
system drift may result in reproductive incompatibility between isolated populations,
even in the absence of any sort of adaptive, or environmental change. 
Does this neutral system drift matter, i.e., lead to measurable consequences?
To address this,
we use quantitative genetic theory to estimate how quickly 
reproductive incompatibility due to system drift manifests.
